Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-31 _tex/paper2.tex
Built site for gh-pages

In the sample-comparison model, we assume a more realistic scenario than
the conceptual-population model. Specifically, we assume researchers
have access to a data sample consisting of a \emph{limited number of
judgments \((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a
specific number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\)}, drawn from the conceptual population. In a subsequent
step, we assume that judges \emph{do not make all possible comparisons}
allowed by the sample, but rather only \(n_{C}\) comparisons.

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the data sampling
process using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:

@@

\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if the data for the } i \text{ element of } I \text{ is observed} \\
0 & \text{if the data for the } i \text{ element of } I \text{ is missing}
\end{cases}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mat23}{
ia_{(ia)} = \begin{cases} 
1 & \text{if the data for the } i,a \text{ elements of } IA \text{ are observed} \\
0 & \text{if the data for the } i,a \text{ elements of } IA \text{ are missing}
\end{cases}
}\end{equation}

@@

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). The sum of its elements represents the total data sample
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals from the population.

@@

The DAG also depicts the \emph{sample design} vector \(S\) as a causal
factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
matrix (see equation (\ref{eq-mat12})), that is, this vector determines
\emph{which repeated comparisons the subset of judges makes for the
subset of stimuli produced by the subset of individuals} sampled from
the conceptual-population. This process ultimately generates the
sample-comparison outcome \(O^{sc}_{R}\) from the conceptual-population
outcome \(O^{cp}_{R}\).

Notably, the DAG shows that \(S\) is independent of all other variables
in the model. This implies that the conceptual model represented in
DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
judgment, judge, stimulus, and individual has the same probability of
being included in the sample as any other observation within their
respective groups \citep[pp.~16-112]{Lawson_2015}.

@@

\[
c_i = \begin{cases} 
1 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is observed} \\
0 & \text{if the data for the } i^{th} \text{ element of } O^{sp}_{R} \text{ is missing}
\end{cases}
\]

@@

when you talk about the variability, refer that the equal dispersion is
just assuming 1 for identification.

@@

\[
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & := f_{e}(X_{I}) \\
  e_{J} & := f_{e}(Z_{J}) \\
  e_{IA} & \:\bot\:e_{JK}
\end{aligned}
\]

>>>>

The sample-comparison model presents a more realistic scenario than the
conceptual-population model. Expressly, we assume researchers work with
a data sample consisting of a \emph{limited number of judgments
\((n^{s}_{K})\) from a sample of judges \((n^{s}_{J})\) and a specific
number of texts \((n^{s}_{A})\) from a sample of students
\((n^{s}_{I})\)}, all drawn from the conceptual population. In a
subsequent step, we assume that judges do not perform \emph{all possible
repeated comparisons} within the data sample. Instead, they conduct a
sufficient number of comparisons, \(n_{C}\), to ensure an accurate
estimation of the proportion \(P(B>A)\).

@@

To incorporate the sampling mechanism and facilitate the interpretation
of the sample-comparison model, we first define the \emph{data sampling
process} using the binary vector variables \(S_{I}\), \(S_{J}\),
\(S_{IA}\), and \(S_{JK}\) as follows:

@@

\begin{equation}\phantomsection\label{eq-mat22}{
i_{(i)} = \begin{cases} 
1 & \text{if data element } i \text{ from } I \text{ is observed} \\
0 & \text{if data element } i \text{ from } I \text{ is missing}
\end{cases}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-mat23}{
ia_{(ia)} = \begin{cases} 
1 & \text{if data elements } i,a \text{ from } IA \text{ are observed} \\
0 & \text{if data elements } i,a \text{ from } IA \text{ are missing}
\end{cases}
}\end{equation}

@@

Finally, we define the \emph{sample design} vector \(S\) in equation
(\ref{eq-mat25}), which maps each element of \(S_{IA}\) to every element
of \(S_{JK}\). Thus, the vector contains \(n\) elements, matching the
number of rows in the \(R\) matrix, and the sum of its elements
represents the total data sample:
\(n^{s} = \binom{n^{s}_{I} \cdot n^{s}_{A}}{2} \cdot n^{s}_{J} \cdot n^{s}_{K}\).
Here, the term \({n^{s}_{I} \cdot n^{s}_{A} \choose 2}\) represents the
binomial coefficient, which quantifies the total number of unique
comparisons possible between every pair of sampled stimuli generated by
each pair of sampled individuals.

@@

The DAG also depicts the \emph{sample design} vector \(S\) as a causal
factor influencing the sample-comparison outcome \(O^{sc}_{R}\). A
square encloses this vector, indicating that it is a conditioned
variable. In this context, \emph{conditioning} means restricting the
statistical analysis to a specific subset of the population based on the
values of a conditioned variable \citep[pp.~32]{Neal_2020}, in this case
\(S\). Essentially, \(S\) serves as the sample equivalent for the \(R\)
matrix (see equation (\ref{eq-mat12})), that is, this vector selects
\emph{all repeated comparisons a subset of judges makes for the subset
of stimuli produced by the sampled individuals}. This process ultimately
generates the sample-comparison outcome \(O^{sc}_{R}\) from the
conceptual-population outcome \(O^{cp}_{R}\).

Notably, the DAG shows that \(S\) is independent of all other variables
in the model. This implies that the conceptual model represented in
DAG~\ref{fig-cj12_dag} applies exclusively to Simple Random Sampling
(SRS) designs \citep[pp.~150]{Kohler_et_al_2019}. In these designs, each
repeated judgment, judge, stimulus, and individual has the same
probability of being included in the sample as any other observation
within their respective groups \citep[pp.~16-112]{Lawson_2015}.

@@

As in the previous section, we first define the \emph{comparison design}
using the binary vector variable \(C\) to facilitate the interpretation
of the sample-comparison model and integrate the comparison mechanism.

\begin{equation}\phantomsection\label{eq-mat26}{
C = \begin{bmatrix}
c_{(1)} \\
\vdots \\
c_{(c)} \\
\vdots \\
c_{(n)} \\
\end{bmatrix}
}\end{equation}

Where each element of \(C\) is a binary value indicating the presence or
absence of data rows in the \(R\) matrix (see equation
(\ref{eq-mat12})):

\[
c_{c} = \begin{cases} 
1 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are observed} \\
0 & \text{if data elements } h, i, a, b, j, k \text{ from } R \text{ are missing}
\end{cases}
\]

Thus, the \emph{comparison design} vector \(C\) contains \(n\) elements
corresponding to the number of rows in the \(R\) matrix, and the sum of
its elements represents the selected comparisons, \(n_{C}\). This
definition assumes that judges do not perform \emph{all possible
repeated comparisons} within the data sample. Instead, they conduct a
sufficient number of comparisons, \(n_{C}\), to ensure the accurate
estimation of the proportion \(P(B>A)\) for each pair of stimuli
\citep[pp.~267]{Thurstone_1927b}.

With the definition of \(C\), the DAG~\ref{fig-cj14_dag} incorporates
the sample and comparison mechanism into the conceptual-population
model. The DAG now also represents the sample-comparison outcome
\(O^{sc}_{R}\) as unobserved, highlighting that researchers cannot
directly access this outcome due to the comparison mechanism. The DAG
also represents the \emph{comparison design} vector \(C\) as a causal
factor influencing the final comparative outcome \(O_{R}\). Equivalent
to the sampling mechanism, a square encloses this vector, indicating
that \(C\) determines \emph{which repeated comparisons the subset of
judges makes for the subset of stimuli produced by the sampled
individuals}. This process ultimately derives the final comparative
outcome \(O_{R}\) from the sample-comparison outcome \(O^{sc}_{R}\).

Notably, the DAG shows that \(C\) is also independent of all other
variables in the model. This independence implies that the conceptual
model represented by DAG~\ref{fig-cj14_dag} applies exclusively to
Random Allocation Comparative Designs \citep[pp.~8]{Bramley_2015} or
Incomplete Block Designs \citep[pp.~259-271]{Lawson_2015}. In these
designs, every repeated judgment has an equal probability of being
included in the sample.

In summary, extending Thurstone's general form with causal inference
allows researchers to address BTL modeling issues, such as judge biases
(see Section~\ref{sec-theory-issue1b}), the hierarchical structure of
stimuli (see Section~\ref{sec-theory-issue1b}), and measurement errors
in trait estimation and hypothesis testing (see
Section~\ref{sec-theory-issue2}). However, the model does not address
concerns about the assumption of equal dispersion discussed in
Section~\ref{sec-theory-issue1a}. Since this issue involves statistical
assumptions regarding the discriminal process distribution, it requires
a statistical model, which we will develop in the next section.

@@

@@

\[
\begin{aligned}
  O_{R} & := f_{C}(O^{sc}_{R}, C) \\
  O^{sc}_{R} & := f_{S}(O^{cp}_{R}, S) \\
  O^{cp}_{R} & := f_{O}(D_{R}) \\
  D_{R} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) \\
  e_{I} & := f_{e}(X_{I}) \\
  e_{J} & := f_{e}(Z_{J}) \\
  e_{IA} & \:\bot\:e_{JK} \\ 
  \\ 
\end{aligned}
\]
